---
title: "Irish Trade Analysis"
author: "Ujwal Mojidra"
format: 
  pdf:
    documentclass: report
    geometry: "top=1.5cm, bottom=1.2cm, left=1cm, right=1cm"
    fontsize: 10pt
    mainfont: "Times New Roman"
    linkcolor: blue
    toc: true  
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{\leftmark}
  - \fancyhead[R]{\thepage}
  - \setlength{\headheight}{15pt}
  - \usepackage{caption}
  - \captionsetup[table]{position=above, skip=10pt}
execute:
  error: true
  message: false
  warning: false
editor: visual
---

## Setup chunk (libraries + options)

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
```

# Phase 1:

## Load & Clean Data

```{r}
trade_raw <- read_csv("raw.csv")
#glimpse(trade_raw)
```

At this point, our columns are:

-   `Statistic Label`
-   `Month`
-   `Commodity Group`
-   `UNIT`
-   `VALUE`

We’ll **never modify `trade_raw` directly** again.

## Clean and standardise variables

```{r}
trade_clean <- trade_raw %>%
  rename(
    statistic = `Statistic Label`,
    month_raw = Month,
    commodity_raw = `Commodity Group`,
    unit = UNIT,
    value = VALUE
  ) %>%
  mutate(
    trade_type = case_when(
      str_detect(statistic, "Imports") ~ "Import",
      str_detect(statistic, "Exports") ~ "Export"
    ),
    month = as.Date(paste0("01 ", month_raw), format = "%d %Y %B"),
    year = lubridate::year(month),
    commodity_code = str_extract(commodity_raw, "\\(([^)]+)\\)") %>%
      str_remove_all("[()]"),
    commodity_desc = str_remove(commodity_raw, "\\s*\\([^)]*\\)"),
    hierarchy_level = if_else(
      nchar(commodity_code) == 1,
      "top",
      "detail"
    )
  ) %>%
  select(
    trade_type,
    month,
    year,
    commodity_code,
    commodity_desc,
    hierarchy_level,
    value
  )
```

What this restructuring block does (important)

-   Converts `"1972 January"` → real `Date`

-   Separates **Import vs Export**

-   Extracts: - `commodity_code` → `0`, `01`, `54`, etc.

-   `commodity_desc` → clean names

-   Tags hierarchy explicitly (`top` vs `detail`)

This is **correct handling** of SITC-style data.

## Split datasets (no double counting ever)

```{r}
trade_top_level <- trade_clean %>%
  filter(hierarchy_level == "top")

trade_detail_level <- trade_clean %>%
  filter(hierarchy_level == "detail")

n_distinct(trade_top_level$commodity_code)
n_distinct(trade_detail_level$commodity_code)
```

From now on:

-   Macro analysis → `trade_top_level`

-   Sector deep dives → `trade_detail_level`

We will **never mix them**.

## Quick sanity checks

```{r}
range(trade_clean$year)
unique(trade_clean$trade_type)
anyNA(trade_clean$month)
```

We could see:

-   Years from **1972 to 2025**

-   Two trade types only

-   No missing dates

## Where we are now

At this point, we have:

-   Clean, reproducible data

-   Correct hierarchy handling

-   A strong analytical base

This alone is already **better than 80% of public trade analyses**.

------------------------------------------------------------------------

# Phase 2:
